The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-11 script/de/chapters/numbertheory.tex
fixed the index links; resolved more pdfTeX warnings by setting plainpages=false (otherwise page x and 10 have the same hypertarget) and hyperfootnotes=false (hyperref does not work with footnotemark/footnoteref)

Die Zahlentheorie oder Arithmetik (hier wird mehr der Aspekt des Rechnens
mit Zahlen betont) wurde von Carl Friedrich Gauss\footnote{%
  Carl Friedrich Gauss, deutscher Mathematiker und Astronom,
  30.4.1777$-$23.2.1855.
}
\index{Gauss, Carl Friedrich}
als besondere mathematische Disziplin begründet. Zu ihren elementaren
Gegenständen gehören: größter gemeinsamer Teiler\footnote{%
Auf ggT\index{ggT}, englisch gcd (greatest common divisor), geht dieser
Artikel im \hyperlink{NumberTheory_Appendix_A}{Anhang A zu diesem Kapitel} ein.
} (ggT), Kongruenzen (Restklassen), Faktorisierung, Satz von Euler-Fermat und
primitive Wurzeln. Kernbegriff sind jedoch die Primzahlen und ihre
multiplikative Verknüpfung.

@@

Für große Moduli scheint es nicht einfach, die Tabelle der multiplikativen
Inversen zu berechnen (das gilt nur für die in den oberen 
Multiplikationstabellen fett markierten Zahlen). Mit Hilfe des kleinen Satzes
von Fermat\index{Fermat!kleiner Satz} kann man dafür einen einfachen 
Algorithmus aufstellen \cite[S. 80]{nt:Pfleeger1997}. Schnellere Algorithmen
werden z.B. in \cite{nt:Knuth1998} \index{Euklidscher Algorithmus!erweiterter}
beschrieben\footnote{%
Mit dem erweiterten Satz von Euklid \index{ggT}(erweiterter ggT) kann man die
multiplikative Inverse berechnen und die Invertierbarkeit bestimmen
(siehe \hyperlink{NumberTheory_Appendix_A}{Anhang A zu diesem Kapitel}).
Alternativ kann auch die Primitivwurzel\index{Primitivwurzel} genutzt werden. 
}.

@@

Ähnliches gilt, wenn $b$ die Unbekannte ist und die Zahlen $a$ und $c$ bekannt sind. \\
Wenn die Eulerfunktion \index{Eulersche Phi-Funktion} $J(m)$ bekannt ist,
findet man leicht\footnote{%
Siehe \hyperlink{NumberTheory_Appendix_A}{Anhang A zu diesem Kapitel}: der größte gemeinsame Teiler (ggT) von ganzen Zahlen.
} $d$ mit $c*d \equiv 1 {\rm ~(mod~} J(m))$ und erhält mit Satz~\ref{thm-zth-pot} 
$$
   a^d \equiv (b^c)^d \equiv b^{c*d} \equiv b^{c*d~(mod~J(m))} \equiv b^1 \equiv b {\rm ~(mod~} m)
$$
die {\em $c$-te Wurzel} $b$ von $a$. \par

@@

\noindent {\bf Schlüsselerzeugung (Schritt 1 bis 3):} \\
%\vskip 5 pt
{\bf 1.} $p=47, q=79$ $( n= 3713;~ J(n) = (p-1)*(q-1)=3588).$ \\
{\bf 2.} $e=37$ (sollte\footnote{%
                Siehe Fußnote~\ref{foot:Selection-of-e} auf Seite
                \pageref{foot:Selection-of-e}.}
                zwischen 79 und 3587 liegen und muss
                teilerfremd\index{Zahlen!teilerfremd (co-prime)}
                zu $3588$ sein). \\
{\bf 3.} $d=97$ (denn $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3589 
\equiv 1{\rm ~(mod~}3588) \;$)\footnote{%
Wie man $d = 97$ mit Hilfe des erweiterten ggT berechnet, wird im \hyperlink{NumberTheory_Appendix_A}{Anhang A zu diesem Kapitel} gezeigt.
}.

>>>>

Die Zahlentheorie oder Arithmetik (hier wird mehr der Aspekt des Rechnens
mit Zahlen betont) wurde von Carl Friedrich Gauss\footnote{%
  Carl Friedrich Gauss, deutscher Mathematiker und Astronom,
  30.4.1777$-$23.2.1855.
}
\index{Gauss, Carl Friedrich}
als besondere mathematische Disziplin begründet. Zu ihren elementaren
Gegenständen gehören: größter gemeinsamer Teiler\footnote{%
Auf ggT\index{ggT}, englisch gcd (greatest common divisor), geht dieser
Artikel in Anhang~\ref{NumberTheory_Appendix_GCD} ein.
} (ggT), Kongruenzen (Restklassen), Faktorisierung, Satz von Euler-Fermat und
primitive Wurzeln. Kernbegriff sind jedoch die Primzahlen und ihre
multiplikative Verknüpfung.

@@

Für große Moduli scheint es nicht einfach, die Tabelle der multiplikativen
Inversen zu berechnen (das gilt nur für die in den oberen 
Multiplikationstabellen fett markierten Zahlen). Mit Hilfe des kleinen Satzes
von Fermat\index{Fermat!kleiner Satz} kann man dafür einen einfachen 
Algorithmus aufstellen \cite[S. 80]{nt:Pfleeger1997}. Schnellere Algorithmen
werden z.B. in \cite{nt:Knuth1998} \index{Euklidscher Algorithmus!erweiterter}
beschrieben\footnote{%
Mit dem erweiterten Satz von Euklid \index{ggT}(erweiterter ggT) kann man die
multiplikative Inverse berechnen und die Invertierbarkeit bestimmen
(siehe Anhang~\ref{NumberTheory_GCD}).
Alternativ kann auch die Primitivwurzel\index{Primitivwurzel} genutzt werden. 
}.

@@

Ähnliches gilt, wenn $b$ die Unbekannte ist und die Zahlen $a$ und $c$ bekannt sind. \\
Wenn die Eulerfunktion \index{Eulersche Phi-Funktion} $J(m)$ bekannt ist,
findet man leicht\footnote{%
Siehe Anhang~\ref{NumberTheory_Appendix_GCD}: der größte gemeinsame Teiler (ggT) von ganzen Zahlen.
} $d$ mit $c*d \equiv 1 {\rm ~(mod~} J(m))$ und erhält mit Satz~\ref{thm-zth-pot} 
$$
   a^d \equiv (b^c)^d \equiv b^{c*d} \equiv b^{c*d~(mod~J(m))} \equiv b^1 \equiv b {\rm ~(mod~} m)
$$
die {\em $c$-te Wurzel} $b$ von $a$. \par

@@

\noindent {\bf Schlüsselerzeugung (Schritt 1 bis 3):} \\
%\vskip 5 pt
{\bf 1.} $p=47, q=79$ $( n= 3713;~ J(n) = (p-1)*(q-1)=3588).$ \\
{\bf 2.} $e=37$ (sollte\footnote{%
                Siehe Fußnote~\ref{foot:Selection-of-e} auf Seite
                \pageref{foot:Selection-of-e}.}
                zwischen 79 und 3587 liegen und muss
                teilerfremd\index{Zahlen!teilerfremd (co-prime)}
                zu $3588$ sein). \\
{\bf 3.} $d=97$ (denn $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3589 
\equiv 1{\rm ~(mod~}3588) \;$)\footnote{%
Wie man $d = 97$ mit Hilfe des erweiterten ggT berechnet, wird in Anhang~\ref{NumberTheory_Appendix_GCD} gezeigt.
}.

### 2009-08-11 script/de/script-de.tex
fixed the index links; resolved more pdfTeX warnings by setting plainpages=false (otherwise page x and 10 have the same hypertarget) and hyperfootnotes=false (hyperref does not work with footnotemark/footnoteref)

% Formatierung der Fußnoten
\usepackage[flushmargin]{footmisc}
\usepackage{fullpage}
\usepackage[hyphens]{url} % options "hyphens" enables linebreaks in \url{}
% Literaturverzeichnis kapitelweise
\usepackage[labelstoglobalaux]{bibunits} % ohne labelstoglobalaux scheint man thebibilograpy und bibtex nicht mischen zu können
\usepackage{float}
\usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black,bookmarksnumbered=true,pdfpagelabels,hypertexnames=false]{hyperref}
%To reduce the number of hyperlink related 'pdfTeX warning's we use options 
%pdfpagelabels and hypertexnames=false and place hyperref after float.
%Option debug helps finding problems.

@@

% be_2005: a) Auf erste Seite nur den Titel ! -> Newpage eingefügt.
%          b) "aboutcryptool.tex" hier und nicht unten vor "introduction.tex"
%             einfügen, damit es VOR dem Inhaltsverzeichnis steht.
\newpage
\input{chapters/aboutcryptool.tex}

>>>>

% Formatierung der Fußnoten
\usepackage[flushmargin]{footmisc}
\usepackage{fullpage}
\usepackage[hyphens]{url} % options "hyphens" enables linebreaks in \url{}
% Literaturverzeichnis kapitelweise
\usepackage[labelstoglobalaux]{bibunits} % ohne labelstoglobalaux scheint man thebibilograpy und bibtex nicht mischen zu können
\usepackage{float}
\usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black,bookmarksnumbered=true,pdfpagelabels,plainpages=false,hyperfootnotes=false]{hyperref}
%To reduce the number of hyperlink related 'pdfTeX warning's we use options 
%pdfpagelabels, plainpages=false and, hyperfootnotes=false and place hyperref after float.
%Option debug helps finding problems.

@@

% be_2005: a) Auf erste Seite nur den Titel ! -> Newpage eingefügt.
%          b) "aboutcryptool.tex" hier und nicht unten vor "introduction.tex"
%             einfügen, damit es VOR dem Inhaltsverzeichnis steht.
\input{chapters/aboutcryptool.tex}

### 2009-08-11 script/en/chapters/numbertheory.tex
fixed the index links; resolved more pdfTeX warnings by setting plainpages=false (otherwise page x and 10 have the same hypertarget) and hyperfootnotes=false (hyperref does not work with footnotemark/footnoteref)

Number theory or arithmetic (the emphasis here is more on the aspect of 
performing calculations with numbers) was established 
by Carl Friedrich Gauss\footnote{%
  Carl Friedrich Gauss, German mathematician and astronomer,
  Apr 30, 1777$-$Feb 23, 1855.
}
\index{Gauss, Carl Friedrich} as a special mathematical discipline. Its 
elementary features include the greatest common divisor\footnote{This 
article deals with the gcd (greatest common 
divisor)\index{Gcd} in \hyperlink{NumberTheory_Appendix_GCD}{Appendix A of this chapter}.}
(gcd), congruence (remainder classes), factorization, the Euler-Fermat theorem
and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{nt:Pfleeger1997}. Quicker algorithms are described, for instance, in 
\cite{nt:Knuth1998}\footnote{Using Euclid's extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see \hyperlink{NumberTheory_Appendix_GCD}{appendix A of this chapter}). Alternatively, we can also use the primitive roots.}.

@@

The situation is similar if 
$b$ is the unknown variable and we know the values of $a$ and $c$: \\
If we know the Euler function \index{Euler!(phi) function} of $m, J(m)$, then we can easily\footnote{See \hyperlink{NumberTheory_Appendix_GCD}{Appendix A of this chapter}:
the greatest common divisor (gcd) of whole numbers.}\index{Gcd} calculate $d$ with $c*d \equiv 1 {\rm ~(mod~} J(m))$ and 
use theorem~\ref{thm-zth-pot} to obtain:
$$
   a^d \equiv (b^c)^d \equiv b^{c*d} \equiv b^{c*d~(mod~J(m))} \equiv b^1 \equiv b {\rm ~(mod~} m)
$$
the {\em $c$-th root} $b$ of $a$. \par

@@

\noindent {\bf Key generation (steps 1 to 3)}:\\
{\bf 1.} $p=47, q=79$ $( n= 3713; ~J(n) = (p-1)*(q-1)=3588).$ \\
{\bf 2.} $e = 37$ (should\footnote{%
                See footnote~\ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $79$ and $3587$ and must be relatively prime
                to $3588$). \\
{\bf 3.} $d=97$ (since $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3589 \equiv 1{\rm ~(mod~}3588) \;$)\footnote{How
to compute $d = 97$ using the {\em extended} gcd algorithm is shown in \hyperlink{NumberTheory_Appendix_GCD}{appendix A of this chapter}}. \\

>>>>

Number theory or arithmetic (the emphasis here is more on the aspect of 
performing calculations with numbers) was established 
by Carl Friedrich Gauss\footnote{%
  Carl Friedrich Gauss, German mathematician and astronomer,
  Apr 30, 1777$-$Feb 23, 1855.
}
\index{Gauss, Carl Friedrich} as a special mathematical discipline. Its 
elementary features include the greatest common divisor\footnote{This 
article deals with the gcd (greatest common 
divisor)\index{Gcd} in appendix~\ref{NumberTheory_Appendix_GCD}.}
(gcd), congruence (remainder classes), factorization, the Euler-Fermat theorem
and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{nt:Pfleeger1997}. Quicker algorithms are described, for instance, in 
\cite{nt:Knuth1998}\footnote{Using Euclid's extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see appendix~\ref{NumberTheory_Appendix_GCD}). Alternatively, we can also use the primitive roots.}.

@@

The situation is similar if 
$b$ is the unknown variable and we know the values of $a$ and $c$: \\
If we know the Euler function \index{Euler!(phi) function} of $m, J(m)$, then we can easily\footnote{See appendix~\ref{NumberTheory_Appendix_GCD}:
the greatest common divisor (gcd) of whole numbers.}\index{Gcd} calculate $d$ with $c*d \equiv 1 {\rm ~(mod~} J(m))$ and 
use theorem~\ref{thm-zth-pot} to obtain:
$$
   a^d \equiv (b^c)^d \equiv b^{c*d} \equiv b^{c*d~(mod~J(m))} \equiv b^1 \equiv b {\rm ~(mod~} m)
$$
the {\em $c$-th root} $b$ of $a$. \par

@@

\noindent {\bf Key generation (steps 1 to 3)}:\\
{\bf 1.} $p=47, q=79$ $( n= 3713; ~J(n) = (p-1)*(q-1)=3588).$ \\
{\bf 2.} $e = 37$ (should\footnote{%
                See footnote~\ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $79$ and $3587$ and must be relatively prime
                to $3588$). \\
{\bf 3.} $d=97$ (since $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3589 \equiv 1{\rm ~(mod~}3588) \;$)\footnote{How
to compute $d = 97$ using the {\em extended} gcd algorithm is shown in appendix~\ref{NumberTheory_Appendix_GCD}}. \\

### 2009-08-11 script/en/script-en.tex
fixed the index links; resolved more pdfTeX warnings by setting plainpages=false (otherwise page x and 10 have the same hypertarget) and hyperfootnotes=false (hyperref does not work with footnotemark/footnoteref)

\usepackage[flushmargin]{footmisc}
\usepackage{fullpage}
\usepackage[hyphens]{url} % options "hyphens" enables linebreaks in \url{}
% Literaturverzeichnis kapitelweise
\usepackage[labelstoglobalaux]{bibunits} % ohne labelstoglobalaux scheint man thebibilograpy und bibtex nicht mischen zu können
%\bibliographyunit[\section] % bibunits für das ganze Dokument aktivieren
\defaultbibliography{references}
\defaultbibliographystyle{abbrv}
\usepackage{float}
\usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black,bookmarksnumbered=true,pdfpagelabels,hypertexnames=false]{hyperref}
%To reduce the number of hyperlink related 'pdfTeX warning's we use options 
%pdfpagelabels and hypertexnames=false and place hyperref after float.
%Option debug helps finding problems.

@@

% Vergleichbar zur Nutzung von Texten aus der Wikipedia (=GNU Free Documentation Licence, GFDL)
% \vspace{200pt}
% \newpage
\begin{quote}
This is a free document, so the content of the document can
be copied and distributed, also for commercial purposes --- as long as
the authors, title and the CrypTool web site (\url{www.cryptool.org})
are acknowledged. Naturally, citations from the CrypTool script are
possible, as in all other documents.\\
The specific license for this document is the GNU Free Documentation Licence.

@@

\newpage
\input{chapters/aboutcryptool.tex}

>>>>

\usepackage[flushmargin]{footmisc}
\usepackage{fullpage}
\usepackage[hyphens]{url} % options "hyphens" enables linebreaks in \url{}
% Literaturverzeichnis kapitelweise
\usepackage[labelstoglobalaux]{bibunits} % ohne labelstoglobalaux scheint man thebibilograpy und bibtex nicht mischen zu können
%\bibliographyunit[\section] % bibunits für das ganze Dokument aktivieren
\defaultbibliography{references}
\defaultbibliographystyle{abbrv}
\usepackage{float}
\usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black,bookmarksnumbered=true,pdfpagelabels,plainpages=false,hyperfootnotes=false]{hyperref}
%To reduce the number of hyperlink related 'pdfTeX warning's we use options 
%pdfpagelabels, plainpages=false and, hyperfootnotes=false and place hyperref after float.
%Option debug helps finding problems.

@@

% Vergleichbar zur Nutzung von Texten aus der Wikipedia (=GNU Free Documentation Licence, GFDL)
% \vspace{200pt}
\begin{quote}
This is a free document, so the content of the document can
be copied and distributed, also for commercial purposes --- as long as
the authors, title and the CrypTool web site (\url{www.cryptool.org})
are acknowledged. Naturally, citations from the CrypTool script are
possible, as in all other documents.\\
The specific license for this document is the GNU Free Documentation Licence.

@@

\input{chapters/aboutcryptool.tex}
